Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2025-02-14 analysis/paper/paper.qmd
work on the result section

Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the skeletal compositions, fragmentation, and butchery patterns of deer bones were examined. The distribution of body portions and butcher patterns can shed light on how indigenous people utilize deer in response to economic demands during the foreign presence. The proportion of deer skeletal elements was quantified using NISP and classified into groups based on the meat yield: meaty elements (e.g., upper forelimb and upper hindlimb ), mid-meaty elements (e.g., body parts), and less-meaty elements (e.g., skull bones and lower limb) [@Madrigal2004]. While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain. Thus, non-meaty elements were considered as secondary source of nutrition in the interpretation. Similarly, the degree of fragmentation could also imply the extraction of marrow or grease. The degree of fragmentation is calculated by dividing Minimal Number of Elements (MNE) by NISP. For butcher patterns, cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were identified and quantified for a comparison between different temporal sequences. In addition, modifications such as cut marks, fracturing, and burning were examined, especially the cut marks associated with the skinning process. Skinning processes can be identified by the encircling and straight marks left on the lower limb, such as metapodials, which result from the series of cuts typically made to acquire intact animal hides [@Costamagno2019; @Isaakidou2007].

@@

boar_reg <- summary(lm(boar ~ total, data = wider_mammal_count))
boar_lm <- c(round(boar_reg$adj.r.squared, 2), sprintf('%.3f', boar_reg$coefficients[,4][2]))
munt_reg <- summary(lm(muntjac ~ total, data = wider_mammal_count))
munt_lm <- c(round(munt_reg$adj.r.squared, 2), sprintf('%.2f', munt_reg$coefficients[,4][2]))
deer_reg <- summary(lm(deer ~ total, data = wider_mammal_count))
div_reg <- summary(lm(diversity ~ total, data = wider_mammal_count))
div_lm <- c(round(div_reg$adj.r.squared, 2), sprintf('%.2f', div_reg$coefficients[,4][2]))

```

@@

@@

```{r}
#| label: fig-deer-portion
#| echo: false
#| fig-cap: "The frequency of skeletal portions using NISP of deer across cultural layers at Kiwulan"

knitr::include_graphics(here::here("analysis/figures/deer_portion.png"))
```

Since deer remains was the major component of faunal assemblage at Kiwulan according to their high quantity and ubiquity, their body parts were examined to explore any changes in dietary practices and preferences.

>>>>

Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the frequency of skeletal elements and surface modifications such as cutmarks of deer bones were examined to assess butchery and transport strategy. This can also shed light on how indigenous people utilize deer in response to high economic demand when involved in regional trade network. Skeletal elements was quantified using NISP and classified into groups based on meat yield: meaty (e.g., upper forelimb and upper hindlimb ), mid-meaty (e.g., body parts), and less meaty portions (e.g., skull and lower limb) [@Madrigal2004; @Munro2004]. Teeth without associated cranial elements were excluded here to reduce overrepresentation due to their higher preservation compared to othe parts. While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain [@Madrigal2002]. Thus, less meaty elements were considered as secondary source of nutrition in the interpretation. Similarly, the degree of fragmentation could also imply the extraction of marrow or grease. The degree of fragmentation is calculated by dividing Minimal Number of Elements (MNE) by NISP. For butcher patterns, cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were identified and quantified for a comparison between different temporal sequences. In addition, modifications such as cut marks, fracturing, and burning were examined, especially the cut marks associated with the skinning process. Skinning processes can be identified by the encircling and straight marks left on the lower limb, such as metapodials, which result from the series of cuts typically made to acquire intact animal hides [@Costamagno2019; @Isaakidou2007].

@@

boar_reg <- summary(lm(boar ~ total, data = wider_mammal_count))
boar_lm <- c(round(boar_reg$adj.r.squared, 2), sprintf('%.3f', boar_reg$coefficients[,4][2]))
munt_reg <- summary(lm(muntjac ~ total, data = wider_mammal_count))
munt_lm <- c(round(munt_reg$adj.r.squared, 2), sprintf('%.2f', munt_reg$coefficients[,4][2]))
deer_reg <- summary(lm(deer ~ total, data = wider_mammal_count))
div_reg <- summary(lm(diversity ~ total, data = wider_mammal_count))
div_lm <- c(round(div_reg$adj.r.squared, 2), sprintf('%.2f', div_reg$coefficients[,4][2]))
```

@@

# get the bone mineral desity data
deer_element_BMD <- readr::read_csv(here::here("analysis", "data", "derived_data", "deer_BMD.csv"))
cor_sur_BMD <- cor.test(deer_element_BMD$NISP, deer_element_BMD$BMD)
survivor_cor <- round(c(cor_sur_BMD$estimate, cor_sur_BMD$p.value),2)

@@

Since deer remains was the major component of faunal assemblage at Kiwulan according to their high quantity and ubiquity, the frequency of skeletal elements were examined to explore any changes in processing decisions of meat portions between different phases. Deer bones were grouped into five body portions based on their nutritional value [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), and lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus, and phalanges). The result (@fig-deer-portion: A) shows that the meaty portion, including upper forelimb and upper hindlimb, especially upper forelimb, is the most frequent part throughout the cultural layers, followed by the less-meaty portions, such as lowerlimb and foot. The consistency of higher numbers of upperlimb and lowerlimb suggests that those elements should be transported together (e.g., @Stiner1991), or marrows from upperlimb were commonly utilized. It should be noted that the category of axial such as ribs were generally low across different phases, which might be caused by highly fragmentation, butchery practices or dietary preferences. A significant change is observed from the frequency of head bones that shows a few in CL6, increased and reach the highest number in CL4, and then decreased since then.

Considering varying frequencies of skeletal parts might be mediated by differential bone density, the relationship between the survivorship of deer skeletal elements and bone mineral density was tested by Pearson correlation. The values of bone survivorship was collected from @Lyman1994's research on deer, and calculated by dividing NISP by expected NISP, followed by multiplying 100%. The result (@fig-deer-portion: B) shows that there is no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between skeletal part survivorship and bone density, suggesting that density-mediated bias has little impact on the taphonomic processes of deer bones.

```{r}
#| label: fig-deer-portion
#| echo: false
#| fig-cap: "A: The frequency of skeletal parts of deer using NISP across cultural layers at Kiwulan. B: Bone minerial density agaist %survivorship of deer remains."

knitr::include_graphics(here::here("analysis/figures/deer_elements.png"))
```
